---
title: "M08-Reflection Essay-Advanced Data Types"
subtitle: "Revealjs Presentation"
author: "Raymond Marquez"
institute: "GBA 5910, Cal Poly Pomona"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: serif
    width: 1600
    height: 900
    footer: Advanced Data Types
    transition: slide
    transition-speed: default
    incremental: false
    toc: true
    toc-depth: 1
    slide-number: true
    scrollable: true
    smaller: true
    code-fold: false
    code-overflow: wrap
    number-sections: false
    number-depth: 5
    embed-resources: true
    css: styles.css
editor: visual
execute: 
  echo: true
  freeze: auto
---

## In Step 1, Dr. Dave Robinson demonstrated how you can deal with various data types using string, forecast, and lubricate packages. Pick three functions he used and how it helped him achieve his wrangling goals.

In this session, the focus was on combining different tools to manage messy and mixed data types efficiently. Three functions that stood out to me were `str_detect()`, `ymd()`, and `auto.arima()`.

1.   **`str_detect()`** was helpful in searching for patterns within text. It allows filtering rows based on whether a specific word or pattern appears, which is useful when dealing with unstructured text data.

2.   **`ymd()`** simplified the conversion of character strings into proper date objects. This is important because working with dates requires a consistent format for sorting and calculations.

3.   **`auto.arima()`** from the forecast package was used to automatically fit a time series model. This function made it possible to move from cleaned data to forecasting without complex manual setup.

These tools showed how different types of data can be handled smoothly within one workflow.

## Summarize the presentation by Dr. Pat Scholoss, a Professor at the University of Michigan, on how to deal with date data in Step 2. How much were you able to understand his presentation? Compare his style with Dave Robinson's. What do you like or dislike?

This presentation focused on handling date-time data correctly and the challenges that come with it. One of the key points was that date formats are often inconsistent and must be standardized before analysis.

Some important ideas included:

1.   Converting dates into a consistent format

2.   Extracting useful components like year and month

3.   Avoiding errors that come from incorrect parsing

I was able to understand the general concepts, especially the importance of cleaning date data early. However, some sections required more attention because they were more detailed.

## In Step 3, I introduced the common usage of gtsummary, which can be useful for school work. Summarize the package's capabilities as explained. How may the {gtsummary} benefit you for your school work or career?

The `{gtsummary}` package is designed to create well-structured summary tables directly in R.

Its capabilities include:

1.   Producing descriptive statistics

2.   Comparing groups

3.   Presenting regression results

4.   Formatting tables automatically

This package is especially useful for academic work because it reduces the need to manually format tables. It also improves the overall presentation of results. For future work, it allows quick and professional reporting of findings, which is important when communicating results to others.

## In Step 4, you watched the extended presentation of the {gtsummary} package by Dr. Daniel Dsjoberg, the very creator of the package. How does it differ from gt and gtExtras? Give three things you learned newly that were not explained in the lecture in Step 3.

In this extended session, the differences between `{gtsummary}`, `gt`, and `gtExtras` became clearer.

1.   `{gtsummary}` focuses on statistical summaries and analysis outputs

2.   `gt` is mainly used to build and structure tables

3.   `gtExtras` adds visual styling and enhancements

Three additional things I learned were;

> Regression models can be turned into formatted tables with very little code
>
> Statistical tests such as p-values can be added automatically
>
> Tables can be exported easily into formats like Word and PDF

These features make `{gtsummary}` more powerful for reporting and sharing results.

## In Step 5, you explored my advanced codebook. Pick one new interesting function from each of the three topics -- strings, factor, and date-time -- that was not introduced in M07. Describe what they are and what they can do for you. While doing so, use actual codes and outputs.

### String Function: `str_trim()`

```{r}

library(stringr)

text <- "   data wrangling   "
str_trim(text)
```

This function removes extra spaces from the beginning and end of a string, which helps clean messy text data.

### Factor Function: `fct_infreq()`

```{r}

library(forcats)

categories <- factor(c("A", "B", "A", "C", "A", "B"))
fct_infreq(categories)
```

This function orders factor levels based on how often they appear, which is useful when summarizing or plotting categorical data.

### Date-Time Function: `month()`

```{r}

library(lubridate)

date <- ymd("2024-04-01")
month(date, label = TRUE)
```

This extracts the month from a date and can display it as a label, making it useful for grouping data.

## I will repeat this again here. Data wrangling is like laying a foundation/basement for a building, while visualization is building a structure above the basement. working in a basement is not a glamorous job at all. No one notices it; people will see what they can see above the ground and judge it, but without the hard work you did in the basement, the building will never stand well. Wrangling with strings and date-time data is particularly hard because we don't get to use it as often as we use numeric data. However, you will get to use factors often, so get used to them soon. What did you like the most about working with the tools you learned in this module? Elaborate on your point.

This comparison reflects my experience well. Initially, I paid more attention to visualization, but I came to understand that proper data preparation is what makes those results reliable.

What I appreciated most:

-    Turning messy data into a clean format

-    Learning how to handle categorical variables effectively

-    Understanding how small transformations improve analysis

Working with strings and dates required more effort because they are less familiar, but they are very important in real datasets. Factors were more straightforward and useful in organizing results.

This module helped me realize that the quality of any analysis depends heavily on the preparation stage.

## By this time, I hope you found a way to deal with your challenges well. If you still have a problem and want to consult with me, you can come to see me during my office hours. How much time have you been able to study for this module? Which steps have you been able to study thoroughly?

> I spent a reasonable amount of time on the module, focusing more on practical coding than theory. I have a strong understanding of Steps 1 and 3, a moderate grasp of Step 4, and found Step 2 (date-time handling) the most challenging.
>
> My main difficulties were working with date-time data, revisiting complex concepts, and managing time across all sections. Moving forward, I plan to practice more with `lubridate`, work with real datasets, and continue using `{gtsummary}` to improve my skills.
