---
title: "M05-2-Application: Data Wrangling with Tidyverse in R"
author: "Raymond Marquez"
date: '`r format(Sys.time())`'
format: 
  html: 
    toc: true
    toc-depth: 4
    number-sections: true
    code-line-numbers: true
    code-fold: false
    code-link: true
    embed-resources: true
editor: visual
execute: 
  freeze: auto
  warning: false
  error: true
---

# [Exercise 1: airquality Data Management]{.smallcaps}

## [Section 1]{.smallcaps}

```{r}

library(tidyverse)

# Convert to tibble 
air.tib <- as_tibble(airquality)

# Convert back to data frame 
air.df <- as.data.frame(air.tib)

class(air.df) 

```

## [Section 2]{.smallcaps}

```{r}

# Tidyverse approach
air.tib %>%
  filter(Temp == 97) %>%
  select(Ozone)

# Base R alternative using subsetting
air.df[air.df$Temp == 97, "Ozone", drop = FALSE]
```

# [Exercise 2: nycflights13 - Flight Delays Analysis]{.smallcaps}

## [Section 1]{.smallcaps}

```{r}

library(nycflights13)

monthly_extremes <- flights %>%
  filter(!is.na(arr_delay)) %>%
  group_by(month, day) %>%
  slice_max(arr_delay, n = 10) %>%
  summarise(daily_top_avg = mean(arr_delay), .groups = "drop_last") %>%
  summarise(avg_delay = mean(daily_top_avg))

print(monthly_extremes)


```

## [Section 2]{.smallcaps}

```{r}

ggplot(monthly_extremes, aes(x = factor(month), y = avg_delay)) +
  geom_col(fill = "darkorange2", width = 0.7) +
  labs(title = "Monthly Profile of Extreme Arrival Delays",
       subtitle = "Mean of the 10 longest delays per day",
       x = "Month of Year", 
       y = "Mean Delay (mins)") +
  theme_classic()
```

## [Section 3]{.smallcaps}

[***Conclusion***]{.underline}

*The data indicates that peak arrival delays are remarkably consistent throughout the year. While one might expect spikes during winter storms or summer thunderstorms, the high level delay threshold remains a possibilty regardless of the season.*

# [Exercise 3: Religion and Income Patterns]{.smallcaps}

## [Section 1]{.smallcaps}

```{r}

# Overview of structure
glimpse(relig_income)

# Check for empty cells across the entire set
colSums(is.na(relig_income))
```

***relig_income** data set is a tibble with 18 rows and 11 columns.*

## [Section 2]{.smallcaps}

```{r}

?relig_income 
```

*The data comes from a Pew Research Center survey and has two main variables;*

-   *Religion*

-   *Income categories*

## [Section 3]{.smallcaps}

*We'll pivot the data to make it long using the pivot_long() function, because, currently, the data is in a wide format, making it hard to visualize income distributions all through the different religious affliations.*

## [Section 4]{.smallcaps}

```{r}
library(tidyr)
library(ggplot2)
library(dplyr)

# Data reshaping (identical to original)
relig_long <- relig_income %>%
  pivot_longer(-religion, names_to = "income", values_to = "count") %>%
  filter(income != "Don't know/refused")

# Alternative grouped bar chart using geom_bar(stat = "identity")
ggplot(relig_long, aes(x = religion, y = count, fill = income)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_viridis_d() +
  labs(title = "Income Distribution by Religious Affiliation",
       x = "Religion", y = "Number of Respondents", fill = "Income Bracket") +
  theme_minimal()

```

# [Exercise 4: US Rent and Income Correlation]{.smallcaps}

## [Section 1]{.smallcaps}

```{r}

# Summary 
summary(us_rent_income)

# Quick dimensions and head
dim(us_rent_income)

head(us_rent_income, 10)
```

*The tiblbe has 105 rows and 5 columns.*

## [Section 2]{.smallcaps}

```{r}

?us_rent_income
```

The data was captured in a 2017 American Community Survey

Description of the data variables is as follows:

***GEOID** : FIP state identifier*

***NAME:** Name of state*

***variable:** Variable name: income = median yearly income, rent = median monthly rent*

***estimate:** Estimated value*

***moe:** 90% margin of error*

## [Section 3]{.smallcaps}

> *Currently the data is in a long format, so each states appears two times. We need to move the "variable" column into its own headers to allow for a scatter plot comparison using the **pivot_wider()*** function.

## [Section 4]{.smallcaps}

```{r}

# Widening the data
rent_wide <- us_rent_income %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe))

# Scatter plot with a regression trend
ggplot(rent_wide, aes(x = estimate_income, y = estimate_rent)) +
  geom_point(color = "midnightblue", size = 2, alpha = 0.7) +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE) +
  labs(title = "Income vs. Rent: State-Level Analysis",
       x = "Median Annual Income", 
       y = "Median Monthly Rent") +
  theme_light()
```

## [Section 5]{.smallcaps}

```{r}

# Widening the data
us_rent_income %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe))%>%
 ggplot(aes(x = estimate_income, y = estimate_rent)) +
  geom_point(color = "midnightblue", size = 2, alpha = 0.7) +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE) +
  labs(title = "Income vs. Rent: State-Level Analysis",
       x = "Median Annual Income", 
       y = "Median Monthly Rent") +
  theme_light()
```

# Exercise 5: Sports Data Manipulation

```{r}

mba <- data.frame(
  player = c('A', 'A', 'B', 'B', 'C', 'C'),
  year   = c(1, 2, 1, 2, 1, 2),
  stats  = c('22/2/3', '29/3/4', '18/6/7', '11/1/2', '12/1/1', '19/2/4')
)
mba
```

## [Section 1]{.smallcaps}

```{r}

mba_sep <- mba %>%
  separate(stats, into = c("points", "assists", "steals"), sep = "/", convert = TRUE) #separating

mba_sep

```

## [Section 2]{.smallcaps}

```{r}

mba_uni <- mba_sep %>%
  unite(stats, points, assists, steals, sep = "-") #uniting

mba_uni
```

# [Exercise 6: Table Generation with `gt`]{.smallcaps}

## [Section 1]{.smallcaps}

```{r}

library(gt)

# Finding the daily 'latest start' per month
late_starts <- flights %>%
  drop_na(dep_time) %>%
  group_by(month, day) %>%
  summarise(first_out = min(dep_time), .groups = "drop") %>%
  group_by(month) %>%
  slice_max(first_out, n = 1)

late_starts

# Generate a'fancy' table
late_starts %>%
  gt() %>%
  tab_header(title = "Monthly Peak for First Daily Departure") %>%
  cols_label(month = "Month", day = "Day", first_out = "Time") %>%
  data_color(columns = first_out, palette = "Greens")
```

## [Section 2]{.smallcaps}

```{r}

#head(flights, 5)

top_destinations <- flights %>%
  count(dest, sort = TRUE) %>% #count and sort 
  slice_head(n = 5)

top_destinations %>% 
  gt() %>%  # generate a quality table
  tab_header(title = "Top Five Destinations from New York")
```

The data makes sense since these are major cities that receive a lot of flights from New York City (NYC)

# Exercise 7: Handling Missing Solar Radiation Data

```{r}

# imputing missing values with the median

# calculate the median
solar_median <- round(median(air.tib$Solar.R, na.rm = TRUE))
solar_median

# replacing the NAs with the median
air_clean <- air.tib %>%
  replace_na(list(Solar.R = solar_median))

# Check if NAs are cleared
anyNA(air_clean$Solar.R)
```

# Exercise 8: Temperature Analysis

```{r}

# total and maximum temperatures combined with label recoding
air_summary <- air_clean %>%
  mutate(Month_Label = month.name[Month]) %>% # Built-in R month names
  group_by(Month_Label) %>%
  summarise(
    Total_Heat = sum(Temp),
    Peak_Heat = max(Temp)
  )

head(air_summary)
```

# Exercise 9: Temperature Analysis - Visual

```{r}

# visualizing Max Temp
ggplot(air_summary, aes(x = Peak_Heat, y = reorder(Month_Label, Peak_Heat))) +
  geom_segment(aes(x = 60, xend = Peak_Heat, yend = Month_Label), color = "grey") +
  geom_point(size = 4, color = "darkred") +
  labs(title = "Highest Recorded Temperature by Month", x = "Temp (°F)", y = "") +
  theme_minimal()
```

# Exercise 10: Multi-Variable Ozone Comparison

```{r}

air_clean %>%
  mutate(T_W_Ratio = Temp / Wind) %>%
  pivot_longer(cols = c(Temp, Wind, T_W_Ratio), names_to = "Metric", values_to = "Val") %>%
  ggplot(aes(x = Val, y = Ozone)) +
  geom_point(color = "darkcyan", alpha = 0.4) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +
  facet_wrap(~Metric, scales = "free_x") +
  labs(title = "Environmental Drivers of Ozone Levels") +
  theme_bw()
```

[***Interpretation***]{.underline}

*Ozone levels show a strong positive correlation with heat. Interestingly, wind acts as a diffuser, higher wind speeds are consistently linked to lower Ozone concentrations. The ratio provides a combined metric, but the individual variables offer clearer predictive power.*
